BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/results.tex
removed the todo items
removed a question mark
spell check and added some more stats

\chapter{Simulation}
\label{sec:SimulationResults}
Two levels of simulation are discussed in this section; first of single voxels,
and second of a single slice (64x64 voxels).  The single time series tests
investigate the ability of particle filters to estimate parameters of the BOLD model 
in a noisy environment. Single voxel tests were
performed eleven different times, each with a new noise realization.
Repeating tests with different noise realizations demonstrates the 
particle filter's resiliance to noise and explores the variance of
the model.

@@

The following tests all use less data (shorter sequences), on par with 
the average length of an FMRI session. For this low noise case 
$(\sigma_y = 0.001,\ \sigma_x = .0005)$, the eleven realizations are shown in 
\autoref{fig:LowNoiseRealization}.
The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
Overall, \autoref{fig:PreprocessedLowNoise} shows that the preprocessing was 
effective at removing trends, although the spline did struggle to match the signal 
torward the end. This slight drift effect was caused by the final median being 
well above the base
level. 
At several time points the particle filter successfully filtered the input
(\autoref{fig:FitComparisonLowNoise}).
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm had converged sufficiently to prevent such inexplicable movement.
A similar circumstance occurs at around 100 seconds in. A combination of 
noise and preprocessing biased the results toward a peak signal above the true peaks. 
The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}.

@@

Note that region 4 has a very low $\epsilon$. For this reason, the only areas
with significant estimates of the BOLD time series were 1,2 and 3. With
such a low $\epsilon$, region 4 was below the noise threshold. Notice that the
regions 1, 2 an 3 stick out in both the residual and the mutual information
map, indicating that the particle filter was successful in matching those regions. 
Mutual information was an extremely succesfull metric, with the exception of 
a few false positives. Thus, another heatmap (\autoref{fig:sim_hm2}) is 
helpful to show what threshold would remove those false positives.

\begin{figure}
\centering
\subfigure{
\includegraphics[width=10cm]{images/sim_hm_15_6.png}}
\subfigure{\includegraphics[width=1cm]{images/scale6.png}}
\caption{More stringent mutual information heatmap. Higher (yellow) is better.}
\label{fig:SimSNRhm}
\end{figure}

@@

%todo quantify the number of false positives, add extra Heatmap removing false positives
The histograms again demonstrate that a single point estimate of the parameters 
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation; which is usually
performed with statistical parametric maps. The thresholds applied to this
slice, both for mutual information and residuals is arbitrary, and needs further
research. Tighter thresholds removes the false positives present
in the images at the cost of false negatives. As noted in \autoref{sec:SingleVoxelReview}
the false positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument from
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in the 
\autoref{fig:simslice_hm}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This is why there are holes in regions
2 and 3. These results indicate that the particle filter is effective
at regressing against a noisy signal.

>>>>

\chapter{Simulation}
\label{sec:SimulationResults}
Two levels of simulation are discussed in this section; first of single voxels,
and second of a single slice (64x64 voxels).  The single time series tests
investigate the ability of particle filters to estimate parameters of the BOLD model 
in a noisy environment. Single voxel tests were
performed eleven different times, each with a new noise realization.
Repeating tests with different noise realizations demonstrates the 
particle filter's resilience to noise and explores the variance of
the model.

@@

The following tests all use less data (shorter sequences), on par with 
the average length of an FMRI session. For this low noise case 
$(\sigma_y = 0.001,\ \sigma_x = .0005)$, the eleven realizations are shown in 
\autoref{fig:LowNoiseRealization}.
The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
Overall, \autoref{fig:PreprocessedLowNoise} shows that the preprocessing was 
effective at removing trends, although the spline did struggle to match the signal 
toward the end. This slight drift effect was caused by the final median being 
well above the base
level. 
At several time points the particle filter successfully filtered the input
(\autoref{fig:FitComparisonLowNoise}).
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm had converged sufficiently to prevent such inexplicable movement.
A similar circumstance occurs at around 100 seconds in. A combination of 
noise and preprocessing biased the results toward a peak signal above the true peaks. 
The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}.

@@

Note that region 4 has a very low $\epsilon$. For this reason, the only areas
with significant estimates of the BOLD time series were 1,2 and 3. With
such a low $\epsilon$, region 4 was below the noise threshold. Notice that the
regions 1, 2 an 3 stick out in both the residual and the mutual information
map, indicating that the particle filter was successful in matching those regions. 
Mutual information was an extremely successful metric, with the exception of 
a few false positives. Thus, another heatmap (\autoref{fig:sim_hm2}) is 
helpful to show what threshold would remove those false positives.

\begin{figure}
\centering
\subfigure{
\includegraphics[width=10cm]{images/sim_hm_15_6.png}}
\subfigure{\includegraphics[width=1cm]{images/scale6.png}}
\caption{More stringent mutual information heatmap. Higher (yellow) is better.}
\label{fig:sim_hm2}
\end{figure}

@@

There were 1479 voxels analyzed, and there 11 were false positives, according
the mutual information threshold of $0.1$. Raising the threshold to $0.15$ reduced
the number of false positives to $0$. Quantifying false negatives is harder, because
of the presence of white matter.

The histograms again demonstrate that a single point estimate of the parameters 
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation; which is usually
performed with statistical parametric maps. The thresholds applied to this
slice, both for mutual information and residuals is arbitrary, and needs further
research. Tighter thresholds removes the false positives present
in the images at the cost of false negatives. As noted in \autoref{sec:SingleVoxelReview}
the false positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument from
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in the 
\autoref{fig:simslice_hm}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This is why there are holes in regions
2 and 3. These results indicate that the particle filter is effective
at regressing against a noisy signal.
